Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/elementary.tex
fixed some missing hyphens in cache types
done spell check. some errors found

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar
\cc{sim-cache}, simulating an 8KB data cache with a 32-byte cache line and
separate instruction cache. Results shown are for a direct mapped cache.]
{\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}.]
{\label{elementary branches}\includegraphics[scale=\myscale]{plots/on2_-_Branches.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Cache and branch prediction simulation results for elementary sorts}
\label{Cache and branch prediction simulation results for elementary sorts}
\end{figure}
\newpage

@@

The branch behaviour of bubble sort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubble sort finds the smallest (or largest) key in the
unsorted part of the array and moves it to start (end) of the array.
Thus, we would expect that bubbble sort's branch prediction behaviour
would be similar to that of selection sort.

However, in the proces of moving the smallest element into position,
bubble sort also rearranges other array elements. Selection sort
sweeps across the array and always keeps track of the left-to-right
minimum up to the current point. However, bubble sort actually moves
this minimum, exchanging pairwise with each element it encounters
which is larger. Thus, even the unsorted part of the array becomes
partially sorted. We might expect that bubble sort's comparison branch
for a partially sorted array might be more predictable than for a
random one, but it is not. With a random array, the smaller elements
are equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubble sort, the last place that we look. Thus,
we change the candidate minimum much more often than if we were
searching a randomly ordered array.

>>>>

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar
\cc{sim-cache}, simulating an 8KB data cache with a 32-byte cache line and
separate instruction cache. Results shown are for a direct-mapped cache.]
{\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}.]
{\label{elementary branches}\includegraphics[scale=\myscale]{plots/on2_-_Branches.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Cache and branch prediction simulation results for elementary sorts}
\label{Cache and branch prediction simulation results for elementary sorts}
\end{figure}
\newpage

@@

The branch behaviour of bubble sort is particularly interesting when
compared to selection sort. Although we usually think of the two
algorithms as being different, they are actually very similar. Like
selection sort, bubble sort finds the smallest (or largest) key in the
unsorted part of the array and moves it to start (end) of the array.
Thus, we would expect that bubble sort's branch prediction behaviour
would be similar to that of selection sort.

However, in the process of moving the smallest element into position,
bubble sort also rearranges other array elements. Selection sort
sweeps across the array and always keeps track of the left-to-right
minimum up to the current point. However, bubble sort actually moves
this minimum, exchanging pairwise with each element it encounters
which is larger. Thus, even the unsorted part of the array becomes
partially sorted. We might expect that bubble sort's comparison branch
for a partially sorted array might be more predictable than for a
random one, but it is not. With a random array, the smaller elements
are equally likely to appear anywhere in the array. With a partially
sorted array, the smaller elements tend to appear towards the start of
the array which is, in bubble sort, the last place that we look. Thus,
we change the candidate minimum much more often than if we were
searching a randomly ordered array.

### 2005-08-14 docs/tech_report/merge.tex
done spell check. some errors found

Base mergesort has the same number of mispredictions as Knuth's sorts (see
Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
`\n{Right}' refer to interating through the first (or left-hand) and second (or
right-hand) lists being compared, respectively.

@@

This predilication is likely to be due to the fact that the final merge goes
forwards. The left merge step is also favoured, but this is merely due to the
fact that it occurs first in the code. While there are more left branches, the
actual number of branches taken are the same.

>>>>

Base mergesort has the same number of mispredictions as Knuth's sorts (see
Figure \ref{mergesorts-3}, indicating a possible minimum number. `\n{Left}' and
`\n{Right}' refer to iterating through the first (or left-hand) and second (or
right-hand) lists being compared, respectively.

@@

This predilection is likely to be due to the fact that the final merge goes
forwards. The left merge step is also favoured, but this is merely due to the
fact that it occurs first in the code. While there are more left branches, the
actual number of branches taken are the same.

### 2005-08-14 docs/tech_report/quick.tex
done spell check. some errors found

Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
the indiviual branches in base quicksort. The \n{i} branch is the loop control
branch for the index sweeping from left-to-right up the array during partition.
The \n{j} branch is the corresponding loop sweeping down the array. The
partition step terminates when \n{i $\geq$ j}, and the \n{Partition End} branch
checks that condition. The \n{Insertion} branch is the comparison branch in the
insertion sort algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a final pass of
insertion sort over the whole array to sort these small partitions. Finally, the
\n{Median} branch is the sum of the results for the branches involved in
computing the median-of-three for each partition.

@@

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \n{i} loop will
execute a large number of times, and thus the \n{i} loop will be biased towards
being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
because the average element is likely to be smaller than the pivot.  Therefore,
the \n{j} loop will execute only a small number of times. Thus, we see that in a
particular partition, we do not normally get an equal number of executions of
the \n{i} and \n{j} loops. Instead we get a large number of executions in one
loop (which makes its branch relatively predictable), whereas the other loop
often does not enter at all (which makes its branch also predictable, although
executed a much smaller number of times). Thus, although overall the \n{i} and
\n{j} branches are taken twice as often as not-taken, on a particular partition
one will usually be biased in the not-taken direction, and as a result will be
executed a relativelty small number of times.

@@

Figure \ref{quick_median_cycles} shows the cycle count of each of the
quicksorts, showing that a pseduo-median-of-seven offers the greatest tradeoff
between the reduced number of branches and instructions, and the increase in
branch mispredictions.

@@

Finally, in multi-quicksort, it may be possible to choose better pivots.
Choosing better (ie, more evenly distributed pivots) would reduce the number of
containers. The advantage of this is that it reduces linearly the number of
pivots to be searched by the sequential search. The most this can be reduced by
is a factor of three, indicating perfectly distributed pivots. This would reduce
the instruction overhead of the multi-quicksort considerably.

>>>>

Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
the individual branches in base quicksort. The \n{i} branch is the loop control
branch for the index sweeping from left-to-right up the array during partition.
The \n{j} branch is the corresponding loop sweeping down the array. The
partition step terminates when \n{i $\geq$ j}, and the \n{Partition End} branch
checks that condition. The \n{Insertion} branch is the comparison branch in the
insertion sort algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a final pass of
insertion sort over the whole array to sort these small partitions. Finally, the
\n{Median} branch is the sum of the results for the branches involved in
computing the median-of-three for each partition.

@@

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \n{i} loop will
execute a large number of times, and thus the \n{i} loop will be biased towards
being taken. The \n{j} loop, on the other hand, will be biased toward not-taken,
because the average element is likely to be smaller than the pivot.  Therefore,
the \n{j} loop will execute only a small number of times. Thus, we see that in a
particular partition, we do not normally get an equal number of executions of
the \n{i} and \n{j} loops. Instead we get a large number of executions in one
loop (which makes its branch relatively predictable), whereas the other loop
often does not enter at all (which makes its branch also predictable, although
executed a much smaller number of times). Thus, although overall the \n{i} and
\n{j} branches are taken twice as often as not-taken, on a particular partition
one will usually be biased in the not-taken direction, and as a result will be
executed a relatively small number of times.

@@

Figure \ref{quick_median_cycles} shows the cycle count of each of the
quicksorts, showing that a pseudo-median-of-seven offers the greatest trade-off
between the reduced number of branches and instructions, and the increase in
branch mispredictions.

@@

Finally, in multi-quicksort, it may be possible to choose better pivots.
Choosing better (i.e., more evenly distributed pivots) would reduce the number of
containers. The advantage of this is that it reduces linearly the number of
pivots to be searched by the sequential search. The most this can be reduced by
is a factor of three, indicating perfectly distributed pivots. This would reduce
the instruction overhead of the multi-quicksort considerably.
